Examining background gene lists in clusterProfiler - RNA-seq example

### 2024-05-31 analysis/analyse_rnaseq2.Rmd
effect of gene list length

# gs = gene sets
# gs format is list of symbols
orabad <- function( df, gs ) {

@@

  defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
  defup <- defup[lapply(strsplit(defup," "),length)==2]
  defup <- unique(sapply(strsplit(defup," "),"[[",2))

  if ( length(defup)<250 ) {
    defup <- head(rownames(subset(def,log2FoldChange>0)),250)
    defup <- defup[lapply(strsplit(defup," "),length)==2]
    defup <- unique(sapply(strsplit(defup," "),"[[",2))
  }

  defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
  defdn <- defdn[lapply(strsplit(defdn," "),length)==2]
  defdn <- unique(sapply(strsplit(defdn," "),"[[",2))

  if ( length(defdn)<250 ) {
    defdn <- head(rownames(subset(def,log2FoldChange<0)),250)
    defdn <- defdn[lapply(strsplit(defdn," "),length)==2]
    defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
  }

@@

oragood <- function( df, gs ) {

@@

@@

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]])
  good <- oragood(d[[1]],gs[[i]])
  compare_ora_plots(bad,good)
} )

>>>>

# gs = gene sets
# gs format is list of symbols
orabad <- function( df, gs, ngenes=2000 ) {

@@

  defup <- head(rownames(subset(def,log2FoldChange>0)),ngenes)
  defup <- defup[lapply(strsplit(defup," "),length)==2]
  defup <- unique(sapply(strsplit(defup," "),"[[",2))

  defdn <- head(rownames(subset(def,log2FoldChange<0)),ngenes)
  defdn <- defdn[lapply(strsplit(defdn," "),length)==2]
  defdn <- unique(sapply(strsplit(defdn," "),"[[",2))

@@

oragood <- function( df, gs, ngenes=2000 ) {

@@

Now run an exhaustive analysis, changing the gene list length from 125 to 2000.

```{r,run2}

ngenes <- c(125,250,500,1000,2000)
params <- expand.grid(1:length(d),1:length(gs),ngenes)
params2 <- expand.grid(names(d),names(gs),ngenes)

colnames(params) <- c("d","gs","ngenes")

ora_dat <- mclapply(1:nrow(params), function(i) {
  j=params[i,1]
  k=params[i,2]
  l=params[i,3]
  bad <- orabad(d[[j]],gs[[k]],l)
  good <- oragood(d[[j]],gs[[k]],l)
  res <- compare_ora(bad,good)
  return(res)
} , mc.cores=8)

lapply(ngenes, function(n) {
  res2 <- subset(res1,Var3==n) ; summary(res2$nsig_bad)
} )

lapply(ngenes, function(n) {
  res2 <- subset(res1,Var3==n) ; summary(res2$nsig_good)
} )

lapply(ngenes, function(n) {
  res2 <- subset(res1,Var3==n) ; summary(res2$jaccard)
} )

# NUMBER OF SIGNIFICANT SETS

ns1 <- lapply(ngenes ,function(n) {
  res2 <- subset(res1,Var3==n)
  ns <- matrix( res2$nsig_bad , ncol=9 )
  rownames(ns) <- names(d)
  colnames(ns) <- names(gs)
  return(ns)
} )

names(ns1) <- ngenes

ns1

ns1 <- lapply(ns1,function(y) {
  res <- apply(y,2,function(x) {
    x <- x[!is.na(x)]
    x <- x[is.finite(x)]
    mean(x)
  } )
  return(res)
})

ns1

ns2 <- lapply(ngenes ,function(n) {
  res2 <- subset(res1,Var3==n)
  ns <- matrix( res2$nsig_good , ncol=9 )
  rownames(ns) <- names(d)
  colnames(ns) <- names(gs)
  return(ns)
} )

names(ns2) <- ngenes

ns2

ns2 <- lapply(ns2,function(y) {
  res <- apply(y,2,function(x) {
    x <- x[!is.na(x)]
    x <- x[is.finite(x)]
    mean(x)
  } )
  return(res)
})

ns2

# JACCARD
jac <- lapply(ngenes ,function(n) {
  res2 <- subset(res1,Var3==n)
  ns <- matrix( res2$jaccard , ncol=9 )
  rownames(ns) <- names(d)
  colnames(ns) <- names(gs)
  return(ns)
} )
names(jac) <- ngenes

jac

jac <- lapply(jac,function(y) {
  res <- apply(y,2,function(x) {
    x <- x[!is.na(x)]
    x <- x[is.finite(x)]
    mean(x)
  } )
  return(res)
})

jac

jac <- do.call(rbind,jac)

plot(jac[,"c5.go"],type="b", pch=19,cex=1.5,lwd=2,
  xaxt = "n",ylim=c(0,1),ylab="jaccard",xlab="no. genes selected")
axis(1, at = seq(1, 5, by = 1),
     labels = ngenes)
points(jac[,"c2.cp.reactome"],type="b",col="red",pch=19,cex=1.5,lwd=2)
points(jac[,"c8.all"],type="b",col="blue",pch=19,cex=1.5,lwd=2)
points(jac[,"c3.tft.gtrd"],type="b",col="gray50",pch=19,cex=1.5,lwd=2)
mtext("impact of gene list length on severity of FDR problem")

legend("bottomright", legend=c("GO","Reactome", "Markers", "TFT"),
       col=c("black", "red", "blue", "gray50"), lty=1, cex=1.2,lwd=3,pch=19)

@@

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=125)
  good <- oragood(d[[1]],gs[[i]],ngenes=125)
  compare_ora_plots(bad,good)
} )

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=250)
  good <- oragood(d[[1]],gs[[i]],ngenes=250)
  compare_ora_plots(bad,good)
} )

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=500)
  good <- oragood(d[[1]],gs[[i]],ngenes=500)
  compare_ora_plots(bad,good)
} )

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=1000)
  good <- oragood(d[[1]],gs[[i]],ngenes=1000)
  compare_ora_plots(bad,good)
} )

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=2000)
  good <- oragood(d[[1]],gs[[i]],ngenes=2000)
  compare_ora_plots(bad,good)
} )
